---
title: "MMR Vaccination Coverage and the 2026 U.S. Measles Resurgence"
author: "Elizabeth Meehan, MPH<br>`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                       fig.width = 9, fig.height = 6.5)
library(tidyverse)
library(ggrepel)
library(scales)
```

## Introduction

Measles cases in the United States surged in 2026, nearly matching the
full-year 2025 total within the first six months of the year. This report
investigates a central hypothesis behind that resurgence: declining
MMR (measles-mumps-rubella) vaccination coverage among kindergartners,
which has fallen nationally from 95.2% (2019-20) to 92.5% (2024-25) —
below the ~95% threshold generally considered necessary for herd immunity.

**Research question:** How has MMR vaccination coverage changed in recent
years, and how does it relate to 2026 measles case counts across states?

This report consolidates the analysis built across five R scripts
(`01_clean_nndss_measles.R` through `05_pa_county_detail.R`). Full
data-cleaning detail, known pitfalls, and data-source caveats are
documented in the project `README.md`; this report focuses on the
analysis and findings, reading from the already-cleaned files in
`data/clean/`.

## Data Sources

- **CDC NNDSS Weekly Data** — weekly measles case counts (Imported +
  Indigenous) by state, 2022-2026.
- **JHU CSSE county-level MMR dataset** — 2-dose MMR vaccination coverage
  by county, school years 2017-18 through 2024-25.

Three different "national measles case count" figures circulate for 2026
(NNDSS, CDC's dedicated measles dashboard, and the JHU tracker), which do
not agree exactly due to differing update cadences and case-classification
rules. This report uses NNDSS throughout for consistency; see the README
for the full reconciliation note.

## Methodology Summary

MMR coverage was aggregated from county to state level two ways:

- **Track A (Snapshot):** each state's most recent available school year
  of county data, averaged (unweighted) to a single current-ish coverage
  estimate. Covers 46 of 51 states/DC (Arkansas and West Virginia have no
  usable county-level data).
- **Track B (Trend):** restricted to counties with data in at least 5 of
  8 school years, to track the same counties over time. Covers 35 states.

Three state/year combinations (Louisiana 2020-21 & 2021-22, Missouri
2019-20, Utah 2017-18) were explicitly excluded as methodology-mismatched
per the source data dictionary, rather than silently coerced to missing.
This report uses **Track A** throughout, since it answers the
cross-sectional "coverage vs. current outbreak" question central to this
analysis.

---

## Finding 1: MMR Coverage Trend Over Time

```{r load-trend-data}
mmr_trend <- read_csv("data/clean/mmr_state_trend_20260704.csv")
```

```{r trend-chart, fig.height=6.5}
highlight_states <- c("SC", "PA", "TX", "UT")

track_b_avg <- mmr_trend %>%
  group_by(school_year) %>%
  summarise(mean_mmr_rate = mean(mean_mmr_rate, na.rm = TRUE), .groups = "drop") %>%
  mutate(State = "Track B average (35 states)")

plot_data <- mmr_trend %>%
  filter(State %in% highlight_states) %>%
  select(State, school_year, mean_mmr_rate) %>%
  bind_rows(track_b_avg) %>%
  mutate(
    school_year = factor(school_year, levels = c(
      "2017-18","2018-19","2019-20","2020-21",
      "2021-22","2022-23","2023-24","2024-25"
    )),
    line_group = if_else(State == "Track B average (35 states)", "Average", "Highlighted state")
  )

ggplot(plot_data, aes(x = school_year, y = mean_mmr_rate, group = State,
                       color = State, linewidth = line_group)) +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "gray40", linewidth = 0.5) +
  geom_line(na.rm = TRUE) +
  geom_point(size = 1.8, na.rm = TRUE) +
  annotate("text", x = 1, y = 0.955, label = "95% herd immunity threshold",
           hjust = 0, size = 3.2, color = "gray40") +
  scale_x_discrete() +
  scale_linewidth_manual(values = c("Average" = 1.4, "Highlighted state" = 0.8), guide = "none") +
  scale_color_manual(values = c(
    "Track B average (35 states)" = "black",
    "SC" = "firebrick", "PA" = "darkorange2", "TX" = "steelblue4", "UT" = "seagreen4"
  ), labels = c(
    "Track B average (35 states)" = "Track B average (35 states)",
    "SC" = "South Carolina", "PA" = "Pennsylvania", "TX" = "Texas", "UT" = "Utah"
  )) +
  scale_y_continuous(labels = percent, limits = c(0.8, 1)) +
  labs(
    title = "MMR Coverage Over Time, 2017-18 to 2024-25",
    subtitle = "Track B subset (states with data in >=5 of 8 school years); gaps reflect years without usable data",
    x = "School year", y = "2-dose MMR coverage (state average)", color = NULL,
    caption = "Source: JHU CSSE county-level MMR dataset.\nTrack B average is an unweighted mean of 35 states' own averages, not population-weighted."
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

**Interpretation:** the Track B average declines from 0.944 (2019-20, 31
states) to 0.905 (2024-25, 34 states) — roughly a 4-point drop over five
years. More strikingly, **all four outbreak states highlighted in this
report show a consistent, monotonic decline with no rebound**: South
Carolina (0.973 → 0.951), Pennsylvania (0.971 → 0.932), Texas
(0.970 → 0.940), and Utah (0.924 → 0.881). This distinguishes two
different risk profiles: SC, PA, and TX each started **above** the 95%
threshold and declined **below** it over this period, while Utah shows
**chronic below-threshold coverage** that never approached 95% at any
point shown. Both trajectories ended in a major 2026 outbreak, suggesting
there may be more than one path to the same outcome — a state crossing
the threshold from above, or a state that has simply never been near it.

---

## Finding 2: State-Level Coverage vs. 2026 Case Counts

```{r load-state-data}
coverage_vs_cases <- read_csv("data/clean/coverage_vs_cases_joined_20260704.csv")
```

```{r scatterplot}
states_to_label <- coverage_vs_cases %>%
  slice_max(order_by = cases_2026_ytd, n = 8)

ggplot(coverage_vs_cases, aes(x = mean_mmr_rate, y = cases_2026_ytd)) +
  geom_vline(xintercept = 0.95, linetype = "dashed", color = "firebrick", linewidth = 0.6) +
  annotate("text", x = 0.95, y = max(coverage_vs_cases$cases_2026_ytd) * 0.95,
           label = "95% herd immunity threshold", hjust = 1.05, size = 3.2,
           color = "firebrick") +
  geom_point(aes(color = coverage_group), size = 2.5, alpha = 0.8) +
  geom_text_repel(data = states_to_label, aes(label = state_name),
                   size = 3, color = "black", max.overlaps = 20,
                   box.padding = 0.5, seed = 42) +
  scale_color_manual(values = c("Below 95%" = "steelblue4", "At/above 95%" = "gray60")) +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = percent) +
  labs(
    title = "State MMR Coverage vs. 2026 Measles Case Counts",
    subtitle = "Each point is one state; coverage is most recent available county-average (Track A snapshot)",
    x = "Mean 2-dose MMR coverage (most recent available county-average)",
    y = "Cumulative measles cases, 2026 (NNDSS, most recent reported week)",
    color = "Coverage group",
    caption = "Sources: CDC NNDSS Weekly Data; JHU CSSE county-level MMR dataset.\nMMR coverage is an unweighted mean across reporting counties, not population-weighted."
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
```

```{r stats-tests}
cor.test(coverage_vs_cases$mean_mmr_rate, coverage_vs_cases$cases_2026_ytd,
          method = "spearman")

coverage_vs_cases %>%
  group_by(coverage_group) %>%
  summarise(
    n_states = n(),
    median_cases = median(cases_2026_ytd),
    mean_cases = mean(cases_2026_ytd),
    max_cases = max(cases_2026_ytd),
    .groups = "drop"
  )

wilcox.test(cases_2026_ytd ~ coverage_group, data = coverage_vs_cases)
```

**Interpretation:** the Spearman correlation between state-level MMR
coverage and 2026 case counts is essentially zero (rho = 0.014,
p = 0.93), and a grouped comparison (below vs. at/above the 95% threshold)
is also non-significant (Wilcoxon, p = 0.55) — with the "at/above 95%"
group actually showing a *higher* mean case count, driven by a single
outlier state. **A state's statewide average coverage is a weak
predictor of outbreak size on its own.** The scatterplot itself suggests
why: it shows a **threshold/triangle pattern** rather than a smooth
linear one — every large outbreak occurs below 95% coverage, but most
states below 95% still have near-zero cases. Low coverage appears to be
**necessary but not sufficient** for a large outbreak; an actual
introduction (index case) into a susceptible community is also required.
Correlation and rank-sum tests, which assume a roughly monotonic
relationship, aren't well-suited to detecting this kind of gatekeeper
pattern.

---

## Finding 3: Case Study — South Carolina

South Carolina's statewide MMR average (0.951) sits just *above* the 95%
threshold, yet it had the single largest 2026 outbreak in this dataset.
County-level detail explains why.

```{r sc-chart}
sc_counties <- read_csv("data/clean/sc_county_mmr_20260704.csv") %>%
  mutate(County = fct_reorder(County, mmr_rate))

sc_state_avg <- mean(sc_counties$mmr_rate)
n_sc <- nrow(sc_counties)

ggplot(sc_counties, aes(x = mmr_rate, y = County, fill = highlight)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 0.95, linetype = "dashed", color = "black", linewidth = 0.5) +
  geom_vline(xintercept = sc_state_avg, linetype = "dotted", color = "gray30", linewidth = 0.5) +
  annotate("label", x = 1.01, y = n_sc - 1, label = "95% threshold",
           hjust = 0, size = 3, color = "black", label.size = 0, fill = "white") +
  annotate("label", x = 1.01, y = n_sc - 4,
           label = paste0("SC state avg: ", percent(sc_state_avg, accuracy = 0.1)),
           hjust = 0, size = 3, color = "gray30", label.size = 0, fill = "white") +
  scale_fill_manual(values = c(
    "Outbreak epicenter (Spartanburg)" = "firebrick",
    "Lower coverage, no major outbreak" = "goldenrod2",
    "Other SC counties" = "steelblue4"
  )) +
  scale_x_continuous(labels = percent, limits = c(0, 1.16),
                      breaks = c(0, 0.25, 0.5, 0.75, 1.0)) +
  labs(
    title = "South Carolina County-Level MMR Coverage (2024-25)",
    subtitle = "The state average (dotted line) sits just above the 95% threshold,\nbut masks Spartanburg County -- the confirmed epicenter of the 2025-2026 outbreak",
    x = "2-dose MMR coverage", y = NULL, fill = NULL,
    caption = "Source: JHU CSSE county-level MMR dataset (school year 2024-25).\nOutbreak location confirmed via CDC, SC DPH, and NPR reporting."
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom",
        axis.text.y = element_text(size = 7))
```

**Interpretation:** the outbreak was centered in Spartanburg County —
confirmed by CDC scenario assessments, NPR, ABC News, and SC's own
Department of Public Health — where roughly 95% of the state's outbreak
cases occurred. Spartanburg's own 2024-25 MMR rate (0.900) is the
3rd-lowest of SC's 46 counties, closely matching the 88.9% figure
reported independently by SC health officials and multiple news outlets.
This is a concrete, externally-validated example of how an unweighted
statewide average can mask the specific local pocket actually driving an
outbreak.

**Caveat:** Jasper (0.825) and Fairfield (0.894) counties have *lower*
coverage than Spartanburg, yet neither had a major outbreak. Low coverage
creates the conditions for a large outbreak if introduced — it doesn't
mean the lowest-coverage place will necessarily be the one that has one.

---

## Finding 4: Case Study — Pennsylvania

Pennsylvania's 2026 outbreak is a two-county cluster (Lancaster and
Lebanon), which produces a more complicated picture than South Carolina's
single-county epicenter.

```{r pa-chart, fig.height=9}
pa_counties <- read_csv("data/clean/pa_county_mmr_20260704.csv") %>%
  mutate(County = fct_reorder(County, mmr_rate))

pa_state_avg <- mean(pa_counties$mmr_rate)
n_pa <- nrow(pa_counties)

ggplot(pa_counties, aes(x = mmr_rate, y = County, fill = highlight)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 0.95, linetype = "dashed", color = "black", linewidth = 0.5) +
  geom_vline(xintercept = pa_state_avg, linetype = "dotted", color = "gray30", linewidth = 0.5) +
  annotate("label", x = 1.01, y = n_pa - 1, label = "95% threshold",
           hjust = 0, size = 3, color = "black", label.size = 0, fill = "white") +
  annotate("label", x = 1.01, y = n_pa - 4,
           label = paste0("PA state avg: ", percent(pa_state_avg, accuracy = 0.1)),
           hjust = 0, size = 3, color = "gray30", label.size = 0, fill = "white") +
  scale_fill_manual(values = c(
    "Primary epicenter (Lancaster)" = "firebrick",
    "Secondary hotspot (Lebanon)" = "darkorange2",
    "Lower coverage, no major outbreak" = "goldenrod2",
    "Other PA counties" = "steelblue4"
  )) +
  scale_x_continuous(labels = percent, limits = c(0, 1.16),
                      breaks = c(0, 0.25, 0.5, 0.75, 1.0)) +
  labs(
    title = "Pennsylvania County-Level MMR Coverage (2024-25)",
    subtitle = "Lancaster sits clearly below the state average, but Lebanon --\nalso heavily affected by the same outbreak cluster -- does not stand out",
    x = "2-dose MMR coverage", y = NULL, fill = NULL,
    caption = "Source: JHU CSSE county-level MMR dataset (school year 2024-25).\nOutbreak location confirmed via PA Dept. of Health, WHYY, and Philadelphia Inquirer reporting."
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom",
        axis.text.y = element_text(size = 6))
```

**Interpretation:** Lancaster (0.885) is clearly below both the 95%
threshold and PA's state average (0.932) — consistent with the South
Carolina pattern. Lebanon (0.932), however, sits almost exactly at the
state average, despite being heavily affected by the same outbreak
cluster. **This suggests county-level aggregation has the same kind of
blind spot state-level aggregation does, just one level down** — Lebanon's
involvement likely reflects something not visible in a county average at
all (geographic proximity to Lancaster, shared community networks, or a
specific under-vaccinated school or group). The general lesson: whatever
level of geography is used, an outbreak can still be driven by a sub-unit
not visible at that level.

---

## Limitations

- MMR state/county averages are **unweighted** across reporting units —
  not population-weighted — which can pull an average toward small
  rural counties over large urban ones.
- Track A snapshot mixes different "most recent" school years across
  states depending on data availability; not all states' coverage
  figures are from the same year.
- NNDSS case counts may differ from CDC's dedicated measles dashboard and
  the JHU tracker due to different update cadences and classification
  rules (see README).
- Arkansas and West Virginia have no usable county-level MMR data at all.
- This analysis is observational and cross-sectional; it cannot establish
  causation, only association and pattern.

## Conclusion

Statewide MMR coverage alone is a weak predictor of 2026 measles outbreak
size — but that null result is itself informative. The real relationship
appears to be a **threshold/necessary-but-not-sufficient** one: coverage
below ~95% is a precondition for a large outbreak, but not a reliable
predictor of *where* one will occur, since an actual introduction event
is also required. Both case studies examined here (South Carolina,
Pennsylvania) show that the specific driver of an outbreak often lives
at a finer geographic or social grain than the level being averaged over
— a pattern that likely continues below the county level as well.
